---
title: "Motor Vehicle Collisions Decreased in Toronto During and After the Beginning of the COVID-19 Pandemic"
author: "Emily Su"
thanks: "Data and code are available at: https://github.com/moonsdust/toronto-collisions"
date: today
date-format: "D MMMM YYYY"
abstract: "As one of the fastest-growing and densest Canadian cities, road and pedestrian safety are growing concerns in Toronto, especially after the COVID-19 pandemic. This paper looks at trends of collisions from 2017 to 2023 in Toronto neighbourhoods and wards, types of collisions, and the number of pedestrians involved. The results show that motor vehicle collisions and pedestrian involvement in them have decreased during and after the pandemic but are prevalent in the same areas from 2017 to 2023. Further investigation is needed on the demographics of Toronto areas with high numbers of motor vehicle collisions."
format:
  pdf:
    toc: true
number-sections: true
bibliography: references.bib
---
```{r}
#| include: false
#| warning: false
#| message: false
#### Preamble ####
# Purpose: To create graphs and analyze the number of collisions in different Toronto locations and the types of collisions rom 2017 to 2023.
# Author: Emily Su
# Email: em.su@mail.utoronto.ca
# Date: 23 January 2024
# Prerequisites: All the scripts in the scripts folder have been ran starting from 00-install_packages.R.
```

```{r}
#| include: false
#| warning: false
#| message: false

library("tidyverse") # Contains data-related packages
library("knitr") # To make tables
library("opendatatoronto")
library("janitor") # To clean datasets
library("dplyr")
library("ggplot2") # To make graphs
library("sf") # To make maps
```

```{r}
#| include: false # Would run the code but it will have no output and code is not in the document 
#| warning: false # Warnings are not included 
#| message: false # Messages are not included 

# Read in data

# Map Data
the_map <- readRDS('../data/cleaned_city_wards.RDS')
og_cleaned_map_data <- read_csv('../data/cleaned_map_data.csv', show_col_types = FALSE)
cleaned_map_data <-
  og_cleaned_map_data |>
  filter(
    (round(long) != 0 & round(lat) != 0) # Filter out longitude and latitude that are at 0 
  ) 
cleaned_map_data <- st_as_sf(cleaned_map_data, coords = c('long', 'lat'), crs = 4326)

# Collision Data 
cleaned_collisions_data <- read_csv("../data/cleaned_collisions_data.csv", show_col_types = FALSE)

# Ward Data
cleaned_ward_data <- read_csv("../data/cleaned_ward_data.csv", show_col_types = FALSE)
```

# Introduction

As of 2024, Toronto was reported to have one of the worst traffic congestion in North America [@citeWorstCongestion]. As Toronto's population increases over the years, congestion on the road has increased and road and pedestrian safety is a growing concern for many. Some also believe that roads have gotten unsafer in the years after the pandemic. In a 2023 study done by Ipsos surveying Canadians on road safety, they mentioned that concerns about road safety have increased to 65% in 2023 from 59% in 2021 [@citeIpsos]. To reduce traffic collision injuries and fatalities, the city of Toronto created the Vision Zero Safety Plan in 2016, which contains over 50 safety measures with the ultimate goal of eliminating all fatalities on roads [@visionzero]. An ambitious goal yet it raises the question of how effective is this plan across Toronto over the years. We investigated this question and the following: how do motor vehicle collisions compare from the early years of the plan to the present following the outbreak of the COVID-19 pandemic in 2020?

In this paper, collisions trends were analyzed from 2017 to 2023 specifically into motor vehicle collisions in Toronto neighbourhoods and wards, the types of motor vehicle collisions that occurred, and the number of pedestrians involved in them. In our findings, there was a decline in motor vehicle collisions overall from 2020 onwards. However, we saw that the locations where motor vehicle collisions occurred stayed consistent over the years. We also found that the majority of wards that make up a greater proportion of motor vehicle collisions had average household incomes that are below Toronto's overall average household income. Understanding collision trends over the years in Toronto can aid with identifying areas of improvement of current safety plans like Vision Zero and overall helping others feel safer in the city. 

In the rest of this paper, the data section will cover the datasets used, define variables used by our tables and graphs, and briefly explain the data-cleaning process. In the results section, we will reveal tables and graphs made on our datasets and explain what they show. In the discussion section, we will connect back to the real world and explain what the results could mean, the implications of our results, potential areas of improvement for the paper, and suggestions for future works. Finally, the appendix section will include the entire table of some of our results. 

# Data {#sec-data}

The data used by the paper are "City Wards" [@citecitywards], "Ward Profiles (25-Ward Model)" [@citewardprofiles], and "Police Annual Statistical Report - Traffic Collisions" [@citecollisionsdata]  datasets, which were sourced from the City of Toronto's Open Data Portal. These datasets were accessed using the `opendatatoronto` library [@citeopendatatoronto]. Another dataset that could have been used was the "About Motor Vehicle Collisions involving Killed or Seriously Injured Persons" dataset from Open Data Toronto Portal. However, this dataset's most current record was in 2022 and is outdated for our purposes.

The datasets were simulated, cleaned, analyzed, and tested using the `R` programming language [@citeR], `tidyverse` [@citetidyverse], `knitr` [@citeknitr], `janitor`[@citejanitor], `dplyr` [@citedplyr], `ggplot2` [@citeggplot2], and `sf` [@citesf]. The variables created for the datasets used was formed by grouping rows by existing columns and counting the number of rows with certain column variables. However, not all variables from our datasets were used or is visible on the graph. The original datasets was filtered for the years 2017 to 2023 and rows with missing information were omitted. These are the name of the following datasets with variables used for our analysis: cleaned_city_wards_data (`area_short_code`, `geometry`); cleaned_map_data (`year`, `collision_type`, `long`, `lat`, `neighbourhood`, `pedestrian_involved`, `num_of_collisions`, `yearly_collision_num`, `total_collisions_2017_2023`); cleaned_collisions_data (`year`, `collision_type`, `num_of_collisions`, `num_of_pedestrians`); cleaned_ward_data (`ward_num`, `ward_name`, `pop_num`,  `avg_income`). In our dataset, `year` represents a year from 2017 to 2023, inclusive. To view the cleaned datasets created, see @sec-datasetused under the Appendix section for more information. 

## Police Annual Statistical Report - Traffic Collisions
The Toronto Police Services has defined motor vehicle collisions under 4 different categories: Fatal, Personal Injury, Fail to Remain, and Property Damage [@citecollisionsdata]. Fatal means that a person was killed during the vehicle collision, personal injury indicates a person was injured during the collision, fail to remain means that the person left the scene before after information was provided regarding the collision, and property damage indicates that property has been damaged during the collision. The datasets, cleaned_collisions_data and cleaned_map_data, are cleaned versions of Police Annual Statistical Report - Traffic Collisions dataset and contains around 56 observations and 113,129 observations respectively. The 
cleaned_collisions_data datasets contains a summary of the number of collisions and pedestrians involved for each collision type for each year. On the other hand, cleaned_map_data contains more information regarding the neighbourhood where each collisions occurs.

The 4 types are possible values under the variable `collision_type`, which is also named "Type of Collision". Due to the various ways motor vehicle collisions can be categorized and the complexity of it, our analysis looked at motor vehicle collisions that were defined in only 1 of the 4 categories. `num_of_collisions` denotes the number of collisions that occurred in either a neighbhourhood in a specific year (also called "Number of Motor Vehicle Collisions") or the number of collisions of a certain type of a certain year depending on the context (named as "Number of Collisions"). `num_of_pedestrians` represents the number of pedestrians that were hit in a vehicle collision in each collision type per year. `pedestrian_involved` signifies if a pedestrian was hit a collision with a "Yes" or "No". `yearly_collision_num` and `total_collisions_2017_2023` are the count of collisions in a specific year and the total number of collisions from 2017 to 2023 for each Toronto neighbhourhood. Three variables found in the cleaned_map_data `long`, `lat`, and `neighbourhood`, which represents longitude of collision site, latitude of collision site, and the neighbourhood name and number where collision site occurred in. 

## City Wards 
The City Wards dataset, as known as cleaned_city_wards_data, contains that was used to recreate a map of Toronto with its ward divisions. The following variables were used in our analysis: `geometry` represents the coordinates of the ward divisions and `area_short_code` represents the ward number. 

## Ward Profiles (25-Ward Model)
For the dataset regarding Toronto ward profiles known as cleaned_ward_data, the following variables were created: `ward_name` signifies the ward number, `ward_name` indicates the name of the Toronto ward, `pop_num` represents the population size of Toronto wards as of 2020, and `avg_income` denotes the average household income of each ward as of 2020. The dataset was extracted by selecting the rows "Average total income of households in 2020 ($)" and "Total - Age", where "Total - Age" contained the population size of each ward. 

# Results {#sec-result}
## Motor vehicle collisions across Toronto neighbourhoods and wards (2017-2023) {#sec-location}
As of 2024, There are 25 wards and 158 neighbourhoods in the city of Toronto. The distribution of motor vehicle collisions across different neighbourhoods and wards was investigated and the following was obtained:
```{r}
#| label: tbl-neighbhourhoodtable
#| tbl-cap: Toronto neighbourhoods with the most number of motor vehicle collisions from 2017 to 2023
#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
#| warning: false # Warnings are not included 
#| message: false # Messages are not included 
# To create table corresponding to the map

collisions_num_per_neighbhourhood <- 
  # Order the data based on the total number of collisions from 2017 to 2023
  arrange(og_cleaned_map_data, desc(total_collisions_2017_2023)) |>
  # Choose the columns neighbourhood and total_collisions_2017_2023 to include. 
  select(neighbourhood, total_collisions_2017_2023) |>
  # Keep unique rows 
  unique() |>
  # Create new column for the proportion of motor vehicle collisions 
  mutate( 
    proportion =  round(total_collisions_2017_2023 / sum(total_collisions_2017_2023), 2)
  ) |>
  rename(
    "Neighbourhood Name and Number" = neighbourhood,
    "Number of Motor Vehicle Collisions (2017 to 2023)" = total_collisions_2017_2023,
    "Proportion of Motor Vehicle Collisions (2017 to 2023)"  = proportion
  )
  
collisions_num_per_neighbhourhood |>
  head() |>
  kable() 
```

```{r}
#| label: tbl-wardprofilestats
#| tbl-cap: Minimum, quartiles, median, and maximum of population size and average household income across Toronto wards in 2020 
#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
#| warning: false # Warnings are not included 
#| message: false # Messages are not included

cleaned_ward_data |>
  # Choose pop_num and avg_income as columns to show 
  select(pop_num, avg_income) |>
  rename (
    "Population Number" = pop_num, 
    "Average Household Income (CAD)"= avg_income
  ) |>
  summary() |>
  kable()
```
```{r}
#| label: fig-numcollisionmap
#| fig-cap: Number of motor vehicle collisions from 2017 to 2023 across different Toronto wards. 
#| echo: false
#| warning: false
#| message: false
# Code snippet to add points to map is from https://stackoverflow.com/questions/58676661/geom-sf-mapping-points-shapes
# Code snippet to change colour of plots 
# https://www.rdocumentation.org/packages/ggplot2/versions/2.2.1/topics/scale_colour_gradient
# Code snippet to change theme adapated from  
# https://ggplot2.tidyverse.org/reference/theme.html
ggplot() + 
  # Adds in points regarding number of collisions to graph 
  geom_point(data = cleaned_map_data,
    aes(color = num_of_collisions, geometry = geometry), alpha = 0.05,
    stat = "sf_coordinates"
  ) +
  # Creates different graphs for each year 
  facet_wrap(vars(year), ncol = 4) +
  # Change colours of the points
  scale_color_gradient(low = 'lightblue', high = 'darkblue', name = "Number of Collisions") +
  # Overlays a map of Toronto  
  geom_sf(data = the_map, fill = "transparent") +
  # Add ward number labels 
  geom_sf_text(data=the_map, aes(label = area_short_code), color = '#333333', size = 1.5) + 
  # Changes the font of the title and removes the x and y axis 
  theme(plot.title = element_text(size = 9, face = 'bold'), rect = element_blank(), axis.text.x =
        element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), legend.title = 
          element_text(size = 7)) + 
  labs(x = "Longitude",
       y = "Latitude", 
       title="Number of Motor Vehicle Collisions Across Toronto Wards from 2017 to 2023") +
  theme(legend.position = c(0.9, 0.2))
```
@tbl-neighbhourhoodtable reveals that the neighbourhood with the highest number of motor vehicle collisions and the approximately how many motor vehicle collisions from 2017 to 2023 occurred in each neighbourhood. The neighbourhoods with a high number of collisions are the following: West Humber-Clairville (approximately 2%), Wexford/Maryvale (approximately 2%), South Riverdale (approximately 2%), York University Heights (approximately 2%), Clairlea-Birchmount (approximately 2%), Birchmount (approximately 1%), and Etobicoke City Centre (approximately 1%). The full table of @tbl-neighbhourhoodtable can be found here @tbl-neighbhourhoodtablefull in the Appendix section with the rest of the neighbourhoods making up approximately 1% of collisions or less. 

In @tbl-neighbhourhoodtable, West Humber-Clairville has the largest amount of collisions with 2812 motor vehicle collisions from 2017 to 2023. West Humber-Clairville is located in ward 1, which from @fig-numcollisionmap is the region with the darkests shade of blue and has the highest density of points [@citefinder]. Following this is Wexford/Maryvale, which is located in ward 21 with 2075 collisions [@citefinder]. The rest of the neighbourhood are in the following wards based on @citefinder: South Riverdale is in Ward 14, York University Heights is in Ward 7, Clairlea-Birchmount is in Ward 20, and Etobicoke City Centre is in Ward 2. 

@tbl-wardprofilestats reveals that the mean (also known as the average) population size per ward in Toronto is 110452 people with a standard deviation of 10594 people and the mean for the average household income per ward is \$120,096 with a standard deviation of \$33,980.64. It also indicates that the median for the population size per ward is 110095 people and the median for average household income per ward is $107,300. From @tbl-wardprofile, ward 1's average household income is \$95,200 with a population of 115120 people and ward 21's average household income is \$91,500 with a population of 115120 people. For ward 14, 7, 20, and 2, the average households incomes are \$130,800, \$85,700, \$102,200, and \$146,600, respectively and the population sizes are 104555, 111200, 110095, and 117200 people, respectively. 

When comparing the density of collisions in Toronto for each year with @fig-numcollisionmap, the density stays the same from 2017 to 2023 with ward 1 having the largest amount of collisions. According to @tbl-neighbhourhoodtable, @tbl-wardprofilestats, @fig-numcollisionmap, and @citefinder, the average household incomes for 3 out of 4 of the wards of each neighbourhood that make up 2% of motor vehicle collisions are below the mean and median for average income per ward with their population size being higher than the mean population size per ward. 1 ward in the 2% has an average household income above the mean and median for average income per ward and their population size is below the mean and median population size per ward. The wards of the other 2 neighbourhoods from @tbl-neighbhourhoodtable, Clairlea-Birchmount's average household income is below the mean and median with its population size being above or equal to the mean and median. On the other hand, Etobicoke City Centre's average household income is above the mean and median and the population size is above the mean and median. 

## Types of motor vehicle collisions in Toronto (2017 - 2023) {#sec-types}
```{r}
#| label: tbl-collisiontype
#| tbl-cap: Motor vehicle collisions of different types in Toronto from 2017 to 2023
#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
#| warning: false # Warnings are not included 
#| message: false # Messages are not included 

cleaned_collisions_data_table <-
  cleaned_collisions_data |>
  # Group data by year
  group_by(year) |>
  # Obtain unique rows 
  unique() |>
  # Add in new column for proportions
  mutate(
    proportion =  round(num_of_collisions / sum(num_of_collisions), 2)
  ) |>
  # Ungroup the data 
  ungroup() |>
  # Rename columns 
  rename(
    "Year" = year,
    "Motor Vehicle Collision Type" = collision_type,
    "Number of Collisions" = num_of_collisions,
    "Proportion of Motor Vehicle Collisions (Per Year)" = proportion
  ) |>
  # Choosing columns to display for table
  select("Year", "Motor Vehicle Collision Type", "Number of Collisions", "Proportion of Motor Vehicle Collisions (Per Year)")

cleaned_collisions_data_table |>
  # Creating table 
  kable()
```

```{r}
#| label: tbl-collisionstats
#| tbl-cap: Minimum, quartiles, median, and maximum of motor vehicle collisions of different types in Toronto from 2017 to 2023
#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
#| warning: false # Warnings are not included 
#| message: false # Messages are not included 

# Creates table
cleaned_collisions_data_table |>
  select("Number of Collisions") |>
  summary() |>
  kable()
```

```{r}
#| label: fig-collisionhistogram
#| fig-cap:  Number of motor vehicle collisions in Toronto grouped by types from 2017 to 2023
#| echo: false
#| warning: false
#| message: false

# Referenced tutorial notes from summer STA238 2023 to obtain the code portion of geom_bar(stat="identity")
# Code referenced: Alexander (2023) 
# Creating histogram breaking down the types of motor vehicle collisions.
cleaned_collisions_data |>
  ggplot(mapping = aes(x = year, y = num_of_collisions, fill = collision_type)) +
  geom_bar(stat="identity") +
  ggtitle("Number of Motor Vehicle Collisions in Toronto from 2017 to 2023") +
  labs(x = "Year", y = "Number of Collision", fill = "Type of Collisions") +
  theme(legend.position = "bottom", plot.title = element_text(size = 9, face = "bold"), rect = element_blank(), legend.title = element_text(size = 9), legend.text = element_text(size = 7)) +
  scale_fill_viridis_d()
```
From @fig-collisionhistogram, when comparing the overall number of motor vehicle collisions before 2020 and 2020 onwards, there is a overall annual decrease with the number of collisions. The result also reveals that the number of motor vehicle collisions increased from 2017 to 2019 before decreasing in 2020. Collisions involving property damage remains the majority of collision cases over the years. We can see this decrease from 2019 and 2020 from about 79205 collisions to 43297 collisions. According to @tbl-collisionstats and @tbl-collisiontype, the peak number of property damage collisions from 2017 to 2023 was at 60865 in 2019, which made up 77% of the collisions in that year. However property damage collisions in 2020 was up to 79% of collisions in 2020. We can also see that the lowest number of collisions occurred in 2021 and had stayed below 60865 since 2020. @fig-collisionhistogram and @tbl-collisiontype also reveals that 2020 to 2022, the second most common collision type was Personal injury (making up 10% to 12% of collisions in their respective years). In other years, the second most common motor vehicle collisions are Fail to Remain collisions.

\newpage

## Pedestrians involved in a motor vehicle collision (2017-2023) {#sec-pedes}

```{r}
#| label: tbl-pedestrianhistogram
#| tbl-cap: Number of pedestrians involved in motor vehicle collisions in Toronto from 2017 to 2023
#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
#| warning: false # Warnings are not included 
#| message: false # Messages are not included 

# Setting up table 
cleaned_collisions_data_yearly <-
  cleaned_collisions_data |>
  # Groups data by year
  group_by(year) |>
  # Obtains unique rows
  unique() |>
  # Creates new columns for number of pedestrians involved in collisions for each year, gets the number of collisions for each year, and calculates the proportion of pedestrians involved in collisions each year 
  mutate (
    yearly_num_pedestrians = sum(num_of_pedestrians),
    num_of_collisons_year = sum(num_of_collisions),
    proportion = round(yearly_num_pedestrians / num_of_collisons_year, 3)
  ) |>
  ungroup()
  
# Creates table for pedestrians involved in collisions each year 
cleaned_collisions_data_yearly |>
  # Choose the column to display for the graph
  select(year, yearly_num_pedestrians, proportion) |>
  # Filter out non-unique rows 
  unique() |>
  # Rename variables to be clearer for table 
  rename(
    "Year" = year,
    "Number of pedestrians involved (Per Year)" = yearly_num_pedestrians, 
    "Proportion of pedestrians involved (Per Year)" = proportion
  ) |>
  kable()
```
\newpage
```{r}
#| label: fig-pedestrianhistogram
#| fig-cap: Number of pedestrians involved in motor vehicle collisions from 2017 to 2023 in Toronto. 
#| echo: false
#| warning: false
#| message: false

# Code referenced: Alexander (2023) 
# Creates histogram for the number of pedestrians involved in motor vehicle collisions.
cleaned_collisions_data_yearly |>
  ggplot(mapping = aes(x = year, y = yearly_num_pedestrians)) +
  geom_col(fill="lightblue",colour="lightblue") +
  ggtitle("Pedestrians involved in Motor Vehicle Collisions in Toronto from 2017 to 2023") +
  labs(x = "Year", y = "Number of Collisions Involved") +
  theme(plot.title = element_text(size = 9, face = "bold"), rect = element_blank(), legend.title = element_text(size = 9), legend.text = element_text(size = 7)) 
```
From @fig-pedestrianhistogram, we can see that there has been a decline in pedestrian involvement in motor vehicle collisions from 2017 to 2023. Pedestrian involvement was at its highest point in 2017 with 1748 pedestrians where approximately 24% of the total number of collisions in 2017 involved pedestrians. In 2018, 1601 pedestrians were involved in motor vehicle collisions (approximately 21%). With 2019, 2020, and 2021, it was 1545 (approximately 20%), 1055 (approximately 24%), and 990 (approximately 24%), respectively. In 2022, the number of pedestrians increased to 1241 (approximately 22%) but fell to its lowest number of pedestrian in collisions in year with 968 people in 2023 (approximately 20%).

\newpage

# Discussion
From @sec-result, we found that from 2017 to 2023, the number of motor vehicle collisions has decreased during 2020 and onwards, the number of pedestrians involved in these collisions has decreased, and that hotspots for motor vehicle collisions has stayed consistent. We also found that the most common type of collisions involved property damage and the second most common collision type in 2020 to 2022 was Personal injury and for other years it was Fail to Remain. We also observed that there was a decrease in numbers from 2019 to 2020 such as the number of pedestrians involved in collisions in Toronto decreased by approximately 38% and the number of collisions in Toronto decreased by 59%. Our observations also revealed that the number of pedestrians and number of collisions were below the numbers reached in the years prior to 2020, which was when the COVID-19 pandemic started.

Throughout 2020 to 2021, the Ontario Government imposed Stay-at-Home orders, requiring Canadians to stay at home unless they are travelling for essential purposes such as grocery shopping [@citeStayAtHome]. This could have been one of the reasons why there was a decrease in motor vehicle collisions by 59% and 2020 and 2021 having fewer number of accidents. In 2022, the Ontario Government began plans to reopen Ontario in the beginning of the year, which led to more people on the road and outside and therefore could have resulted in the number of collisions in 2022 being higher than in 2020 and 2021 [@citeReopening]. 

Vision Zero Safety Plan was a plan created by the City of Toronto in 2016 to reduce traffic collision injuries and fatalities [@visionzero]. Based on the decreasing number of pedestrians being involved in collisions over the years, it suggests Vision Zero could have had an influence on this trend. However, the number of motor vehicle collisions increased from 2017 to 2019, which means there are other factors in play that could be further investigated by future works. Vision Zero has also faced criticism due to its lack of focus on addressing road and pedestrian safety in low-income areas [@visionzeroinequality]. While we did not directly looked at the relationship of ward income and number of collisions, which could be investigated in future works, we found that the number of crashes across Toronto stayed consistent from 2017 to 2023 based on @fig-numcollisionmap. The neighbourhoods we found impacted with high amounts of vehicle collisions (composing approximately 2% of total collisions) in @tbl-neighbhourhoodtable were mostly in wards that are below the mean average household income based on data from 2020. 

Despite using datasets with over 100,000 observations, the datasets are not 100% accurate. In data entry, human error has occurred. This is seen with the original dataset containing rows that are partially completed with N/A as some of their entries. As mentioned by @citecollisionsdata, in order to ensure privacy, coordinates to the collision sites were offset to the closest road intersection. This means that the neighbourhoods where collisions were recorded at might not be completely accurate. Other errors that may have occurred includes motor vehicles collisions being purposely misrepresented when entered or inaccurate information was entered by mistake.  

The analysis was faced with several limitations as well. The data on Toronto Wards' average incomes and population size was done in 2020, which may not reflect trends of other years. Due to the complexity of motor vehicle collisions and the different they can be categorized, our analysis looked at collisions that were defined in only 1 of the 4 categories (Fatal, Fail to Remain, Property Damage, and Personal Injury). This means that numbers of motor vehicle collisions and number of pedestrians involvement with them are lower than they would be. Since no dataset currently exists that corresponds neighbourhoods with their respective wards, searches for each neighbourhood and their corresponding ward was done manually using the Toronto's Find Your Neighbourhood page [@citefinder].

Future investigations could be conduct on the demographics of Toronto locations with high number of motor vehicle collisions. Other investigations that could be made includes ones regarding the increase in motor vehicle collisions from 2017 to 2019 and the relationship of ward income and number of collisions. Datasets could also be created that maps Toronto neighbourhoods to their respective wards to allow for more in-depth and accurate analysis of motor vehicle collisions in different neighbourhoods and wards. 

\newpage


\appendix

# Appendix {-}


## Datasets Used {#sec-datasetused}
Due to the length of the datasets horizontally, presenting glimpses of the datasets was not possible in the paper. However, the cleaned datasets that was used can be found here: https://github.com/moonsdust/toronto-collisions/tree/main/outputs/data


## Tables from Results {#sec-tablesfromresults}
```{r}
#| label: tbl-wardprofile
#| tbl-cap: Population size and average household income across Toronto wards in 2020
#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
#| warning: false # Warnings are not included 
#| message: false # Messages are not included 

cleaned_ward_data |>
  rename (
    "Ward Number" = ward_num,
    "Ward Name" = ward_name,
    "Population Number" = pop_num, 
    "Average Household Income (CAD)"= avg_income
  ) |>
  kable()
```


```{r}
#| label: tbl-neighbhourhoodtablefull
#| tbl-cap: Toronto Neighbourhoods with the most number of motor vehicle collisions from 2017 to 2023
#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
#| warning: false # Warnings are not included 
#| message: false # Messages are not included 
collisions_num_per_neighbhourhood |>
  kable() 
```
\newpage


# References
